---
title: "Cvx NMF SFS"
author: "Ida Egendal"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Convex NMF vs. AE vs. NMF
This notbook will investigate deconvolutions performances of NMF, cvx nmf and the autoencoder on a small example

```{python}
import numpy as np
import pandas as pd
from sklearn.cluster import KMeans
import matplotlib
import matplotlib.pyplot as plt
from AAUtoSig_init import AAUtoSig, train_AAUtoSig
import torch
import torch.nn.functional as F
from datetime import date
from sklearn.decomposition import NMF
from Cvx_nmf import convex_nmf
```


## CVX NMF

We generate the siganture and exposure matrix using cvx nmf. In Ding et al. they consider a datamatrix with features as rows (same as Pei), which made sense with the Kmeans implementation. 
We implement it with observations as row, making is comparable to the estimation from the autoencoder. This also means that the Kmeans initialization does not make sense.
$$V = XGW  = FW$$

```{python}
def est_cvx(X, rank, n_iter, reverse = False):
    G_cvx, W_cvx = convex_nmf(X, rank, n_iter, init = "kmeans")
    if reverse:
         signatures_cvx = X@W_cvx
         diagonals_cvx = signatures_cvx.sum(axis = 0)
         signatures_cvx = (signatures_cvx@np.diag(1/diagonals_cvx)).T
         exposures_cvx = (np.diag(diagonals_cvx)@G_cvx.T).T
  
    if not reverse:
        exposures_cvx = X@W_cvx
        signatures_cvx = G_cvx
        diagonals_cvx = signatures_cvx.sum(axis = 0)
        exposures_cvx = exposures_cvx@np.diag(diagonals_cvx)
        signatures_cvx = np.diag(1/diagonals_cvx)@(G_cvx.T)
    
    return(exposures_cvx, signatures_cvx, W_cvx)
```

## AE NMF

Signatures and exposures are extracted using the AE with ReLU + PG nn. and hyperparameters set after my gut feeling 

```{python}
def est_AE(X, rank, n_iter, reverse = False):
    n, p = X.shape
    model_AE = AAUtoSig(feature_dim = p, latent_dim = rank)

    optimizer = torch.optim.Adam(params=model_AE.parameters(), lr=1e-3)
    criterion = torch.nn.MSELoss(reduction='sum')

    X_pd = pd.DataFrame(X)
    _,_,_,signatures_AE, exposures_AE,_ = train_AAUtoSig(epochs = n_iter, model = model_AE, x_train = X_pd, x_test = X_pd, criterion = criterion, optimizer = optimizer, batch_size = 8, do_plot = True)

    if not reverse:
        diagonals_AE = signatures_AE.sum(axis = 0)
        exposures_AE = (exposures_AE.T@np.diag(diagonals_AE)).to_numpy()
        signatures_AE = (np.diag(1/diagonals_AE)@(signatures_AE.T)).to_numpy()
        

    if reverse:
        exposures_AE = model_AE.dec1.weight.data.numpy()
        signatures_AE = X@((model_AE.enc1.weight.data).numpy()).T
        signatures_AE = np.maximum(signatures_AE,0)
    
        diagonals_AE = signatures_AE.sum(axis = 0)

        exposures_AE = (np.diag(diagonals_AE)@exposures_AE.T).T
        signatures_AE = (signatures_AE@np.diag(1/diagonals_AE)).T
    W = (model_AE.enc1.weight.data.numpy()).T
    return(exposures_AE, signatures_AE, W)
```

## NMF

NMF with multiplicative updates

```{python}
def est_nmf(X, rank, n_iter):
    model_NMF = NMF(n_components=rank, solver='mu', beta_loss = 2, max_iter=n_iter, init="random")
    exposures_NMF = model_NMF.fit_transform(X)
    signatures_NMF = model_NMF.components_

    diagonals_NMF = signatures_NMF.sum(axis = 1)
    exposures_NMF = exposures_NMF@np.diag(diagonals_NMF)
    signatures_NMF = np.diag(1/diagonals_NMF)@signatures_NMF
    return(exposures_NMF, signatures_NMF)
```


## Load Data
To invesitgate the SFS of NMF and convex NMF we look at 'real' mutational data. The increase in parameters will increase the non-uniqueness.


```{python}
PCAWG = pd.read_csv(r'Q:\AUH-HAEM-FORSK-MutSigDLBCL222\external_data\Alexandrov_2020_synapse\WGS_PCAWG_2018_02_09\WGS_PCAWG.96.csv')
PCAWG.index = [t[0] + '[' + m + ']' + t[2] for (t,m) in zip(PCAWG['Trinucleotide'], PCAWG['Mutation type'])]
PCAWG = PCAWG.drop(['Trinucleotide', 'Mutation type'], axis = 1)

cancers_PCAWG = [c.split('::')[0] for c in PCAWG.columns]
idx = [c  == "Ovary-AdenoCA" for c in cancers_PCAWG]
count_ovary = PCAWG.iloc[:,idx]
Patient_ID = [c.split('::')[1] for c in count_ovary.columns]

count_ovary = count_ovary.T.to_numpy()
```


## Extractions
In cvx nmf and ae, we consider the formulation:
$$V = XGW  = FW$$
and for NMF:
$$V = HW $$

Now, we extract W and G in CVX NMF and AE and H and W in NMF:

```{python}
n_iter = 600
exp_nmf, sig_nmf = est_nmf(count_ovary, rank  = 3, n_iter = n_iter)
exp_cvx, sig_cvx, W_cvx = est_cvx(count_ovary, rank  = 3, reverse = False, n_iter = n_iter)
exp_ae, sig_ae, W_ae = est_AE(count_ovary, rank  = 3, reverse = False, n_iter = n_iter)

```



```{r SFS1}
library(devtools)
library(ggplot2)

library(SFS)
library(reticulate)

V = as.matrix(py$count_ovary, ncol = 96)
total_n = nrow(V)

rownames(V) = py$Patient_ID

sigs_cvx = as.matrix(py$sig_cvx)
W_cvx = as.matrix(py$W_cvx)

dev1 = sampleSFS(P=W_cvx, E = sigs_cvx, maxIter = 10^5, beta = 0.5, check = 1000)
prob.min = dev1$Eminimum
prob.max = dev1$Emaximum

# Converting into right data frame
sig = rep(c(1:3), dev1$totalIter)
mut <- c("C > A","C > G","C > T","T > A","T > C","T > G")
sub = rep(mut, each = 16)
colnames(V) = sub
V = t(V)

dat1 = data.frame(m = factor(rownames(V), levels=unique(rownames(V))), sub = sub, S = t(prob.min))
datmin = reshape(dat1, varying = colnames(dat1)[-c(1,2)], direction = "long", v.names = "min")
dat1 = data.frame(m = factor(rownames(V), levels=unique(rownames(V))),S = t(prob.max))
datmax = reshape(dat1, varying = colnames(dat1)[-1], direction = "long", v.names = "max")
data2 = merge(datmin,datmax, by = c("m","time","id"))
data2$time = factor(data2$time)
data2$xval = rep(1:96,3)

equal_breaks <- function(n = 3, s = 0.05, ...){
  function(x){
    # rescaling
    d <- s * diff(range(x)) / (1+2*s)
    seq(min(x)+d, max(x)-d, length=n)
  }
}

g1 = ggplot(data2) + geom_bar(aes(x = xval, y = max), stat = "identity", width = 0.78, fill = "tomato2")+
  geom_bar(aes(x = xval, y = min), stat = "identity")+
  facet_grid(rows = vars(time), cols = vars(sub), scales = "free", switch = "x")+theme_bw()+
  theme(text = element_text(size=12, face = "bold"), axis.text.x=element_blank(),axis.text.y = element_text(size = 8),axis.ticks = element_blank(), 
        legend.position = "none",strip.text.y.right = element_text(angle = 0, size = 15), panel.spacing.x = unit(0,"line"),
        strip.background.x = element_rect(color="black", fill="white",linetype="blank"),
        strip.text.x = element_text(size = 9))+ 
  ylab("Probability")+xlab("Mutation types")+ggtitle("Signatures")+
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), breaks=equal_breaks(n=3, s=0.2), 
                     expand = c(0.05, 0))
g1


```

```{r SFS2}
sigs_nmf = as.matrix(py$sig_nmf)
exp_nmf = as.matrix(py$exp_nmf)

dev1 = sampleSFS(P=exp_nmf, E = sigs_nmf, maxIter = 10^5, beta = 0.5, check = 1000)
prob.min = dev1$Eminimum
prob.max = dev1$Emaximum

# Converting into right data frame
sig = rep(c(1:3), dev1$totalIter)
mut <- c("C > A","C > G","C > T","T > A","T > C","T > G")
sub = rep(mut, each = 16)

dat1 = data.frame(m = factor(rownames(V), levels=unique(rownames(V))), sub = sub, S = t(prob.min))
datmin = reshape(dat1, varying = colnames(dat1)[-c(1,2)], direction = "long", v.names = "min")
dat1 = data.frame(m = factor(rownames(V), levels=unique(rownames(V))),S = t(prob.max))
datmax = reshape(dat1, varying = colnames(dat1)[-1], direction = "long", v.names = "max")
data2 = merge(datmin,datmax, by = c("m","time","id"))
data2$time = factor(data2$time)
data2$xval = rep(1:96,3)

equal_breaks <- function(n = 3, s = 0.05, ...){
  function(x){
    # rescaling
    d <- s * diff(range(x)) / (1+2*s)
    seq(min(x)+d, max(x)-d, length=n)
  }
}

g1 = ggplot(data2) + geom_bar(aes(x = xval, y = max), stat = "identity", width = 0.78, fill = "tomato2")+
  geom_bar(aes(x = xval, y = min), stat = "identity")+
  facet_grid(rows = vars(time), cols = vars(sub), scales = "free", switch = "x")+theme_bw()+
  theme(text = element_text(size=12, face = "bold"), axis.text.x=element_blank(),axis.text.y = element_text(size = 8),axis.ticks = element_blank(), 
        legend.position = "none",strip.text.y.right = element_text(angle = 0, size = 15), panel.spacing.x = unit(0,"line"),
        strip.background.x = element_rect(color="black", fill="white",linetype="blank"),
        strip.text.x = element_text(size = 9))+ 
  ylab("Probability")+xlab("Mutation types")+ggtitle("Signatures")+
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), breaks=equal_breaks(n=3, s=0.2), 
                     expand = c(0.05, 0))
g1


```

```{r SFS}
sigs_NMF = as.matrix(py$sig_nmf)
exposures_NMF = as.matrix(py$exp_nmf)

sfs_result = sampleSFS(exposures_NMF,sigs_NMF) 
res = samplesToSVD(Presults = sfs_result$P_lastCheckResults, Eresults = sfs_result$E_lastCheckResults, N = 3)

dat.T = data.frame(x = res$E.points[,1], y = -res$E.points[,2])  # Choose PE.points or ET.points

#### PLOT 
ggplot(dat.T, aes(x = x, y = y))+
  geom_point(size = 0.2, col = c(rep("darkgrey", nrow(dat.T)-500), rep("black",500)), shape = 3)+
  labs(x = expression(alpha[1]), y = expression(alpha[2]))+
  theme_bw()



sigs_AE = as.matrix(py$sig_ae)
W_AE = as.matrix(py$W_ae)

sfs_result = sampleSFS(W_AE,sigs_AE) 
res = samplesToSVD(Presults = sfs_result$P_lastCheckResults, Eresults = sfs_result$E_lastCheckResults, N = 3)

dat.T = data.frame(x = res$E.points[,1], y = -res$E.points[,2])  # Choose PE.points or ET.points

#### PLOT 
ggplot(dat.T, aes(x = x, y = y))+
  geom_point(size = 0.2, col = c(rep("darkgrey", nrow(dat.T)-500), rep("black",500)), shape = 3)+
  labs(x = expression(alpha[1]), y = expression(alpha[2]))+
  theme_bw()


```
